---
title: 'Tutorial 1: Intro to Data Types'
author: "Tim Kelso"
date: "20/09/2021"
output: html_document
---
From background reading:

#Background: Coordinate systems
https://www.earthdatascience.org/courses/use-data-open-source-python/intro-vector-data-python/spatial-data-vector-shapefiles/intro-to-coordinate-reference-systems-python/

https://www.earthdatascience.org/courses/use-data-open-source-python/intro-vector-data-python/spatial-data-vector-shapefiles/geographic-vs-projected-coordinate-reference-systems-python/)

##What is a coordinate reference system
To define the location of something you often use a coordinate system. This system consists of an X and a Y value located within a 2 (or more) -dimensional space.

To define the location of objects on the earth, which is round, you need a coordinate system that adapts to the Earth’s shape. When you make maps on paper or on a flat computer screen, you move from a 3-Dimensional space (the globe) to a 2-Dimensional space (your computer screens or a piece of paper). 

The components of the CRS define how the “flattening” of data that exists in a 3-D globe space. The CRS also defines the the coordinate system itself.

##Components of a CRS
-Coordinate System: the X, Y grid upon which your data is overlayed and how you define where a point is located in space.
-Horizontal and vertical units: The units used to define the grid along the x, y (and z) axis.
-Datum: A modeled version of the shape of the earth which defines the origin used to place the coordinate system in space.
-Projection Information: the mathematical equation used to flatten objects that are on a round surface (e.g. the earth) so you can view them on a flat surface (e.g. your computer screens or a paper map).

##Importance of a CRS
If you have data from the same location that are stored in different coordinate reference systems, they will not line up in any GIS or other program unless you have a program like ArcGIS or QGIS that supports projection on the fly. 
Even then, you will want all of your data in the same projection for performing analysis and processing tasks.

##Two types of coordinate systems
Geographic coordinate systems in decimal degrees are helpful when you need to locate places on the Earth. 
However, latitude and longitude locations are not located using uniform measurement units. 
Thus, geographic CRSs are not ideal for measuring distance. 
This is why other projected CRS have been developed.

Geographic coordinate systems: coordinate systems that span the entire globe (e.g. latitude / longitude).
Projected coordinate Systems: coordinate systems that are localized to minimize visual distortion in a particular region (e.g. Robinson, UTM, State Plane)

Each CRS is optimized to best represent the:
- shape and/or
- scale / distance and/or
- area
of features in a dataset. There is not a single CRS that does a great job at optimizing all three elements: shape, distance AND area. Some CRSs are optimized for shape, some are optimized for distance and some are optimized for area. Some CRSs are also optimized for particular regions - for instance the United States, or Europe.

###Geographic CRS
A geographic CRS uses a grid that wraps around the entire globe. This means that each point on the globe is defined using the SAME coordinate system and the same units as defined within that particular geographic CRS.

###Projected CRS
Spatial projection refers to the mathematical calculations performed to flatten the 3D data onto a 2D plane (your computer screen or a paper map). Projecting data from a round surface onto a flat surface, results in visual modifications to the data when plotted on a map. 

The Universal Transverse Mercator (UTM) system is a commonly used projected coordinate reference system. UTM subdivides the globe into zones, numbered 0-60 (equivalent to longitude) and regions (north and south)

From Eubank Tutorial:
#1.0: Setup Instruction
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sp)
library(raster)
library(rgdal)
```

There are three basic steps to creating spatial data by hand:

A. Create geometric objects (points, lines, or polygons)

B. Convert those geometric objects to Spatial* objects (* stands for Points, Lines, or Polygons)
Geometric objects live in an abstract space (the x-y plane). To make them spatial objects, we also need to include information on how those x-y coordinates relate the places in the real world using a Coordinate Reference System (CRS).

C. (Optional:) Add a data frame with attribute data, which will turn your Spatial* object into a Spatial*DataFrame object.


#1.1 SpatialPoints: Your First Spatial* Object!
``` {r 1.1 Make points}
#create a matrix of x-y points
toy.coordinates <- rbind(c(1.5, 2), c(2.5, 2), c(0.5, 0.5), c(1, 0.25), c(1.5, 
    0), c(2, 0), c(2.5, 0), c(3, 0.25), c(3.5, 0.5))

#use SpatialPoints function to convert into spatial data
my.first.points <- SpatialPoints(toy.coordinates)  # ..converted into a spatial object
plot(my.first.points)
```

``` {r 1.1 Getting a summary}
#getting a summary of these points
summary(my.first.points)
coordinates(my.first.points)
```

Unlike a simple geometric object, a SpatialPoints object has the ability to keep track of how the coordinates of its points relate to places in the real world through an associated “Coordinate Reference System” (CRS – the combination of a geographic coordinate system and possibly a projection), which is stored using a code called a proj4string.


``` {r 1.1 Add a coordinate reference system (CRS)}
#see if a projection is defined
is.projected(my.first.points)
# returns `NA` if no geographic coordinate system or projection; 
# returns FALSE if has geographic coordinate system but no projection.

#In this case, while our SpatialPoints object clearly knows what a CRS is, the Spatial object we just created does not have a projection or geographic coordinate system defined. It is ok to plot, but be aware that for many meaningful spatial operations you will need to define a CRS.


#CRS objects can be created by passing the CRS() function to the code associated with a known projection. You can find the codes for most commonly used projections from www.spatialreference.org.

crs.geo <- CRS("+init=EPSG:32633")  # UTM 33N - this is a known projection
proj4string(my.first.points) <- crs.geo  # define projection system of our data
is.projected(my.first.points)

#check 

```